Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-02-05 WSSSPE3/report/wg_appendix_software_credit.tex
Added line breaks. No content change.
Section 5.8, Appendix I and J: small changes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Group Members}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Initial discussions focused on both various mechanisms for, and the
philosophical approach behind, crediting software in scientific papers. These
began with proposals for various ways to credit software (or other research
products including data) that contributed more significantly than a generic
citation, including:
\begin{itemize}

@@

\item Mozilla Science Lab's recently introduced Contributorship Badges for
Science~\cite{Mozilla_badges}, which provide a badge---associated with
ORCID~\cite{orcid}---for author contributions based on the taxonomy
outlined in Project CRediT.

@@

The discussion then evolved into philosophical questions about the importance or
reliance of a particular work on prior science, materials, or software---in
other words, is there a difference between depending on prior scientific
advances and depending on particular software (or experimental equipment)?
Multiple contributors converged on the conclusion that unique capabilities
require some additional credit. The---albeit limited---consensus was that if a
particular study relied on the unique capabilities of software, data, or an
experimental apparatus, then the authors or developers that created this
capability should be credited somehow.

@@

During the first day of WSSSPE3 breakout sessions, the group discussed the
Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
scientific software. That system assigns unique Digital Object Identifiers
(DOIs)---the same system used for scientific publications---to all content
(e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of this system is to track rights and credit for contributors
to entertainment works in order to distribute revenues---similar to the proposed
transitive credit concept.

The group also discussed separating quantitative measures (e.g., number of
citations) from the value of a work in order to give credit, moving towards
qualitative or anecdotal evidence of value. Other topics brought up included a
form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and
using market penetration or adoption rate in a community as a metric, although
it was not clear how this would be measured. Finally, the concept a software
tool's uniqueness or indispensability to a community was mentioned, with value
being characterized by a particular piece of software either offering unique
capabilities or doing something better, faster, or with less computational
requirements than other offerings.

On the second day of WSSSPE3, the working group decided to put aside the
taxonomy of contributions and focus on software citations to ensure developers
receive credit (regardless of contribution). Eventually, once 
software citations are standardized, the goal would be to return to establishing different
roles\slash contributions for this credit. Following this decision, the group
identified two necessary actions to move forward:
\begin{enumerate}

@@

At a minimum, the metadata needed for software citation %(aside from the DOI for
%the software itself) \choinote{DOI is in the list below}
includes:
\begin{itemize}
    \item Name of software,
    \item DOI for software,
    \item Contributors, in the form of names and ORCIDs,
    \item Software dependencies, in the form of DOIs, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file, e.g., as part of
the GitHub repository. The group also discussed similar efforts such as
CodeMeta\footnote{CodeMeta: \url{https://github.com/codemeta/codemeta}}, an
attempt to codify minimal metadata schemes in JSON and XML for scientific
software and code, and implementing transitive credit via
JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
would be stored for closed-source software.

@@

Although far from a standard practice, examples of citing software in
publications can be found in various scientific communities---notably,
representative samples can be found in astronomy~\cite{astronomy_SW_examples}
and biology~\cite{Howison2015}. The group recommended collecting similar
examples from other communities, then developing a software citation principles
document in concert with the FORCE11 Software Citation Working Group (see
\S\ref{SC:plan} for more details), following the model of the FORCE11 Data
Citation Principles document~\cite{DataCitation2014}.

The group also discussed briefly whether software used directly in a
publication---whether to perform simulation, analysis, or as a dependency for
new software developed---should be distinguished from other references due to
the dependence of the study on these research artifacts. Suggestions included a
separate list of citations (with DOIs) for software and other research objects
that serve this sort of ``vital'' role. Similar recommendations were made by the
credit breakout group at WSSSPE2~\cite{WSSSPE2}.

Finally, although a discrete task from software citations themselves,
significant discussion focused on ensuring software citations are indexed in the
same manner as publications, allowing the construction of a corresponding
software citation network. Currently, software releases can receive citable DOIs
via Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these
citations are not processed by indexers such as Web of Science, Scopus, or
Google Scholar. Thus, either in parallel or following the primary task, the
group will need to reach out to these organizations. Initial conversations with
Elsevier\slash Scopus via Michael Taylor during WSSSPE3 clarified that Scopus
is not yet DataCite DOI aware, and also does not yet have an internal identifier
for software or data (but needs\slash plans to add this support). Taylor said
they prefer a ``software article'' with the usual article metadata (e.g.,
authors, citations), and mentioned Zenodo as an example -- this proposal seemed
to align with our group's discussions. Taylor also mentioned another benefit of the
software and associated DOI on GitHub: in addition to a citation, one could obtain
statistics on usage/downloads/forks, which happens to be what
Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

@@

There currently is no standard mechanism for citing software or
receiving credit for software (akin to publications/citations). Software is
eligible for DOI assignment, but DOI metadata fields are not well tuned or
standardized for software (vs.\ publications). Some software providers apply for
DOIs, but this is not widely adopted. Also, there is no mechanism to cite
software dependencies within software.

@@

Following the meeting, Kyle Niemeyer formally requested the merging of software
citation activities with FORCE11-SCWG. Within a month of the meeting, Niemeyer will
organize a virtual meeting of the group and lead the division of
responsibilities for compiling existing practices of software citation and
including software\slash products in promotion and tenure dossiers. Building off
of these efforts, the next major milestone is drafting the Software Citation
Principles document in collaboration with the SCWG, targeted for April 2016.
While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the
efforts of that group towards the Software Citation Principles document, Kyle
will help coordinate contributions from the WSSSPE group members.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Group Members}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% g

@@

Initial discussions mainly focused on both various mechanisms for, and the
philosophical approach behind, crediting software in scientific papers. These
began with proposals for various ways to credit software (or other research
products including data) that contributed more significantly than a generic
citation, including:
\begin{itemize}

@@

\item Mozilla Science Lab has recently introduced Contributorship Badges for
Science~\cite{Mozilla_badges}, which provides a badge---associated with
ORCID~\cite{orcid}---for identifying author contributions in a paper based on the taxonomy
outlined in Project CRediT.

@@

The discussion then evolved into philosophical questions about the importance or
reliance of a particular work on prior science, materials, or software---in
other words, whether there is a difference between depending on prior scientific
advances and depending on certain software (or experimental equipment).
Multiple contributors converged on the conclusion that unique capabilities
require some additional credit. The---albeit limited---consensus was that if a
particular study relied on the unique capabilities of software, data, or an
experimental apparatus, then the authors or developers that created this
capability should be credited somehow.

@@

During the first day of WSSSPE3 breakout sessions, the group discussed and deliberated over the
Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
scientific software. That system assigns unique Digital Object Identifiers
(DOIs)---the same system used for scientific publications---to all content
(e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of the EIDR system is to track rights and credits for contributors
to entertainment works in order to distribute revenues---similar to the proposed
transitive credit concept.

The group also discussed separating quantitative measures (e.g., number of
citations) from the value of a work in order to give credit, moving towards
qualitative or anecdotal evidence of value. Other topics that were brought up included a
form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and
using market penetration or adoption rate in a community as a metric, although
it was not clear how this would be measured. Finally, the concept a software
tool's uniqueness or indispensability to a community was mentioned, with value
being characterized by a particular piece of software either offering unique
capabilities or doing something better, faster, or with less computational
requirements than other offerings.

On the second day of WSSSPE3, the group decided to put aside the
taxonomy of contributions and focus on software citations to ensure developers
receive credit (regardless of contribution). Eventually, once 
software citations are standardized, the goal would be to return to establishing different
roles or contributions for this credit. Following this decision, the group
identified two necessary actions to move forward:
\begin{enumerate}

@@

At a minimum, the metadata required for software citation includes:
\begin{itemize}
    \item Name of software,
    \item DOI for software,
    \item Contributors, in the form of names and ORCIDs,
    \item Software dependencies, in the form of DOIs, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file, e.g., as part of
the GitHub repository. The group also discussed similar efforts such as
CodeMeta\footnote{CodeMeta: \url{https://github.com/codemeta/codemeta}}, an
attempt to codify minimal metadata schemes in JSON and XML for scientific
software and code, and implementing transitive credit via
JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
would be stored for closed-source software.

@@

Although far from a standard practice, examples of citing software in
publications can be found in various scientific communities---notably,
representative samples can be found in astronomy~\cite{astronomy_SW_examples}
and biology~\cite{Howison2015}. The group recommended collecting similar
examples from other communities, and then developing a set of software citation principles
document in concert with the FORCE11 Software Citation Working Group (see
\S\ref{SC:plan} for more details), following the model of the FORCE11 Data
Citation Principles document~\cite{DataCitation2014}.

The group further discussed briefly whether software used directly in a
publication---whether to perform simulation or analysis, or as a dependency for
new software to be developed---should be distinguished from other references due to
the dependence of the study on these research artifacts. Suggestions included a
separate list of citations (with DOIs) for software and other research objects
that serve this sort of ``vital'' role. Similar recommendations were made by the
credit breakout group at WSSSPE2~\cite{WSSSPE2}.

Finally, although a discrete task from software citations,
significant discussion focused on ensuring software citations are indexed in the
same manner as publications, allowing the construction of a corresponding
software citation network. Currently, software releases can receive citable DOIs
via Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these
citations are not processed by indexers such as Web of Science, Scopus, or
Google Scholar. Thus, either in parallel or following the primary task, the
group will need to reach out to these organizations. Initial conversations with
Elsevier\slash Scopus via Michael Taylor during WSSSPE3 clarified that Scopus
is not yet DataCite DOI aware, and also does not yet have an internal identifier
for software or data (but needs\slash plans to add this support). Taylor said
they prefer a ``software article'' with the usual article metadata (e.g.,
authors, citations), and mentioned Zenodo as an example -- this proposal seemed
to align with our group's discussions. Taylor also mentioned another benefit of the
software and associated DOI on GitHub: in addition to a citation, one could obtain
statistics on usage/downloads/forks, which happens to be what
Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

@@

There currently is no standard mechanism for citing software or
receiving credit for software (akin to citations for publications). Software is
eligible for DOI assignment, but DOI metadata fields are not well tuned or
standardized for software (vs.\ publications). Some software providers apply for
DOIs, but this is not widely adopted. Also, there is no mechanism to cite
software dependencies within software.

@@

Following the meeting, Kyle Niemeyer formally requested the merging of software
citation activities with FORCE11-SCWG. Within a month of the meeting, Niemeyer will
organize a virtual meeting of the group and manage the division of
responsibilities for compiling existing practices of software citation and
including software\slash products in promotion and tenure dossiers. Building off
of these efforts, the next major milestone is drafting the Software Citation
Principles document in collaboration with the SCWG, targeted for April 2016.
While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the
efforts of that group towards the Software Citation Principles document, Kyle
will help coordinate contributions from the WSSSPE group members.

### 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
Section 5.8, Appendix I and J: small changes.

While little training focuses specifically on sustainable
software, a variety of training activities increase researcher awareness 
and engagement with software engineering practices and software professionals.
Research Software Engineers are being recognized as critical contributors to
high quality research; the pathway to acquire and master the relevant skills
is not yet clear; equally those skills required by researchers in general are
also not commonly understood nor routinely developed.

>>>>

While little training focuses specifically on sustainable
software, a variety of training activities could increase researcher awareness of
and engagement with software engineering practices and software professionals.
Research Software Engineers are being recognized as critical contributors to
high quality research; the pathway to acquire and master the relevant skills
is not yet clear; equally those skills required by researchers in general are
also not commonly understood nor routinely developed.

### 2016-02-05 WSSSPE3/report/wg_main_software_credit.tex
Section 5.8, Appendix I and J: small changes.

Modern scientific and engineering research relies considerably on software, 
but currently no standard mechanism exists for citing software or receiving 
credit for developing software akin to receiving credit via citations for 
writing papers. Ensuring that developers of such scientific software receive 
credit for their efforts will encourage additional creation and maintenance. 
Standardizing software citations offers one route to establishing such a
citation and credit mechanism. Software is currently eligible for DOI
assignment, but DOI metadata fields are not well tuned for software compared to
publications. Some software providers apply for DOIs but it is still not widely
adopted. Also, there is no mechanism to cite software dependencies within
software in the same way papers cite supporting prior work.

@@

The majority of the remaining discussion focused on standardizing (1) the
metadata necessary for software to be cited and (2) the mechanism for citing
software in publications. In addition, discussions also oriented around the
indexing of software citations necessary for establishing a software citation
network either integrated with the existing paper citation ecosystem or
complementary to it. See Appendix~\ref{sec:appendix_SW_credit} for a more
detailed summary of the working group's discussion on these topics.

@@

The group already merged with the FORCE11 Software Citation Working Group
(SCWG), and their efforts will focus (over the next six to nine months) on
developing a document describing principles for software citation. Following
publication of that document, the working group will focus on outreach to key
groups (e.g., journals, publishers, indexers, professional societies).
Longer-term plans include working with indexers to ensure that software
citations are indexed and pursuing an open\slash community indexer; these
activities may be organized by future FORCE11 working groups.

>>>>

Modern scientific and engineering research often relies considerably on software, 
but currently no standard mechanism exists for citing software or receiving 
credit for developing software akin to receiving credit via citations for 
writing papers. Ensuring that developers of such scientific software receive 
credit for their efforts will encourage additional creation and maintenance. 
Standardizing software citations offers one route to establishing such a
citation and credit mechanism. Software is currently eligible for DOI
assignment, but DOI metadata fields are not well tuned for software compared to
publications. Some software providers apply for DOIs but it is still not widely
adopted. Also, there is no mechanism to cite software dependencies within
software in the same way papers cite supporting prior work.

@@

The majority of the remaining discussion focused on standardizing (1) the
metadata necessary for software to be cited and (2) the mechanism for citing
software in publications. Moreover, discussions also oriented around the
indexing of software citations necessary for establishing a software citation
network either integrated with the existing paper citation ecosystem or
complementary to it. See Appendix~\ref{sec:appendix_SW_credit} for a more
detailed summary of the working group's discussion on these topics.

@@

The group already merged with the FORCE11 Software Citation Working Group
(SCWG), and their efforts will focus (over the next six to nine months) on
developing a document describing principles for software citation. Following the
publication of that document, the group will focus on outreach to key
groups (e.g., journals, publishers, indexers, professional societies).
Longer-term plans include working with indexers to ensure that software
citations are indexed and pursuing an open\slash community indexer; these
activities may be organized by future FORCE11 working groups.
